---
title: "Cyclistic Project"
author: "So Young Kim"
date: "2024-09-12"
output: html_document
---

# Set up the environment by loading necessary packages
```{r setup, include=FALSE}
library(tidyverse)  # Helps wrangle data
library(conflicted)

# Set dplyr::filter and dplyr::lag as the default choices
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

# STEP 1: COLLECT DATA

#Load Divvy datasets (csv files)
```{r}

q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```


# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

#Inspect column names to ensure consistency across datasets
```{r}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

#Rename columns to match the structure of Q1 2020 data for consistency)
```{r}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,start_station_name = from_station_name
                   ,start_station_id = from_station_id
                   ,end_station_name = to_station_name
                   ,end_station_id = to_station_id
                   ,member_casual = usertype
))
```

```{r}

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,start_station_name = from_station_name
                   ,start_station_id = from_station_id
                   ,end_station_name = to_station_name
                   ,end_station_id = to_station_id
                   ,member_casual = usertype
))
```

```{r}
(q2_2019 <- rename(q2_2019
                   ,ride_id = `01 - Rental Details Rental ID`
                   ,rideable_type = `01 - Rental Details Bike ID`
                   ,started_at = `01 - Rental Details Local Start Time`
                   ,ended_at = `01 - Rental Details Local End Time`  
                   ,start_station_name = `03 - Rental Start Station Name` 
                   ,start_station_id = `03 - Rental Start Station ID`
                   ,end_station_name = `02 - Rental End Station Name` 
                   ,end_station_id = `02 - Rental End Station ID`
                   ,member_casual = `User Type`
))
```

#Check the structure of the datasets to ensure consistency
```{r}
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)
```

#ride_id columns in q2_2019, q3_2019, q4_2019 are different type than that of q1_2021
```{r}
str(q2_2019$ride_id)
str(q3_2019$ride_id)
str(q4_2019$ride_id)
str(q1_2020$ride_id)

```

#rideable_type columns in q2_2019, q3_2019, q4_2019 are different than that of q1_2021
```{r}
str(q2_2019$rideable_type)
str(q3_2019$rideable_type)
str(q4_2019$rideable_type)
str(q1_2020$rideable_type)

```

#Convert inconsistent data types for ride_id and rideable_type to character
```{r}
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
```

```{r}
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
```

```{r}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))
```

#Combine all quarterly datasets into a single dataframe
```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

#Remove irrelevant columns that were discontinued after 2020 
```{r}

all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "tripduration", 
            `01 - Rental Details Duration In Seconds Uncapped`,
            `05 - Member Details Member Birthday Year`, 
            `Member Gender`))
```

#inspect that all columns are good
```{r}
colnames(all_trips)
```


# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

#Basic inspection of the combined data
```{r}                                                              
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics


```

#Issues to address:
#1. Standardize the member_casual labels (e.g., "Subscriber" to "member", "Customer" to "casual").
#2. Add additional date-related columns (day, month, year) for aggregation.
#3. Add a ride_length column for ride duration.
#4. Remove invalid rides (e.g., negative ride durations)

#Begin by check how many observations fall under each usertype
```{r}
table(all_trips$member_casual)
```

#Standardize member_casual labels
```{r}

all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

#Verify the reassignment
```{r}
table(all_trips$member_casual)
```

#Add columns for date-related aggregations (day, month, year, day of the week)
```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

#Add a new column ride_length for the duration of each ride
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

#Inspect the structure of the columns
```{r}
str(all_trips)
```

#Convert ride_length to numeric for analysis
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

#Remove invalid entries (e.g., rides with negative durations or station HQ rides)
```{r}

all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

#Basic descriptive statistics for ride_length
```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```

#Condensed summary of ride_length
```{r}
summary(all_trips_v2$ride_length)
```

#Compare ride length between members and casual users
```{r}

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

#Analyze average ride time by day of the week for each user type
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
          
#Reorder the days of the week for better readability
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

#Re-run the analysis with reordered days of the week
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

#Analyze ridership data by user type and weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
        ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

#Visualize the total number of rides by day of the week and user type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at,label=TRUE)) %>%  #label=TRUE returns the weekday in characters
  group_by(member_casual,weekday) %>% 
  summarize(number_of_rides = n(),average_duration=mean(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=number_of_rides,fill=member_casual)) + geom_col(position = "dodge")
```

#Visualize the average ride duration by day of the week and user type
```{r}

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

#Visualize the median ride duration by day of the week and user type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at,label=TRUE)) %>%  
  group_by(member_casual,weekday) %>% 
  summarize(number_of_rides = n(),median_duration=median(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=median_duration,fill=member_casual)) + geom_col(position = "dodge")
```

#Visualize the total number of rides per month by user type
```{r}
all_trips_v2 %>% 
  mutate(month = month(started_at, label=TRUE)) %>%  
  group_by(member_casual,month) %>% 
  summarize(number_of_rides = n(),average_duration=mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x=month,y=number_of_rides,fill=member_casual)) + geom_col(position = "dodge")
```


